### 2011-03-18 extra/sliver_pldi11/preliminaries.tex

The pruning approach that we present in this paper works on Datalog, a general
language that has been used recently to declaratively express static analyses \cite{Whaley2007,bravenboer}.  Normally, one thinks of
the Datalog program as encapsulating the abstraction, but it will be useful for us
to decouple the two.\footnote{The reason is that we want to make theoretical
statements comparing the same Datalog program across different
abstractions.}  We first define Datalog (\refsec{datalog}), which describes the
computation of a query (think more in the direction of concrete semantics).
Then, we focus on the abstraction (\refsec{abstractions}), which only interacts
with the Datalog program via the input tuples.
Throughout this section, we will use \reffig{graphExample} as a running example.

@@

Intuitively, a Datalog program takes a set of input tuples and derives new
tuples.  To formalize this computation, we define the notation of a derivation.

A {\em derivation} (of the query $\xo$) with respect to a set of input tuples
$X$ is a sequence $\bx = (x_1, \dots, x_n)$ such that
\begin{enumerate}
\item [(i)] for each $i = 1, \dots, n$, we have $x_i \in X$; or
there exists a set of indices $J$ such that $(J,i)$ satisfies the following conditions:
$j < i$ for each $j \in J$, and
there is a rule $z \in \sZ$ and an assignment $f$
such that $f(z.t) = x_i$ and
$\{ x_j : j \in J \} = \{ f(s) : s \in z.\bs \}$.
\item [(ii)] $x_n = \xo$; and
\item [(iii)] 
for each $j = 1, \dots, n\!-\!1$,
there exists $J$ such that $j \in J$ and an index $i$
such that $(J,i)$ satisfies the conditions in (i).
\end{enumerate}
Define $\bD(X)$ to be the set of all derivations with respect to the input tuples $X$.

% Explain
Condition (i) says that each tuple in a derivation
should either be given as an input tuple ($x_i \in X$)
or be the result of some rule $z \in \sZ$.
Condition (ii) says that $\xo$ was finally produced via some sequence of rule applications.
Condition (iii) says that the derivation of $x_n$ is minimal, that is, each tuple was
used in some way; as we will see later, this condition is important because it allow us to
define the set of relevant tuples for pruning.

A Datalog program computes $\bD(X)$.
We say that the query $\xo$ is false (proven) if and only if $\bD(X) = \emptyset$.
Although the query is the ultimate quantity of interest,
the Datalog program can be used to provide more information that is useful for pruning.
Specifically, we define $\bP(X)$ to be the subset of the input tuples $X$
which were used in some derivation:
\begin{align}
\label{eqn:Pdef}
\bP(X) \eqdef \{ x \in X : x \in \bx \in \bD(X) \}.
\end{align}
We call $\bP(X)$ the set of {\em relevant input tuples}.  As we will see later,
any tuple not in this set can be safely pruned.
In fact, $\bP(X)$ also tells us whether the query is true or false.
In particular, $\bD(X) = \emptyset$ if and only if $\bP(X) = \emptyset$.
This equivalence suggests that proving and pruning are intimately related;
in some sense, proving the query is just pruning away the query tuple.
In the remainder of the paper, we will make heavy use of $\bP$ as the principal
proving/pruning operation.

@@

We can compute $\bP(X)$ in Datalog by using the Datalog program transformation technique described in 
\cite{liang11minimal}; this allows us to use off-the-shelf Datalog solvers which have already been optimized.
Specifically, we define a set of new relations $\sR' = \{ r' : r \in \sR \}$.
For a term $t = t.r(t.\ba)$ we let $t' = t.r'(t.\ba)$ be the term that using the corresponding new relation.
We then add the following new Datalog rules:
\begin{align}
\xo' &\dlogUpdate \xo. \label{eqn:revBase} \\
s'   &\dlogUpdate z.t', z.\bs. \label{eqn:indBase}
\end{align}
The key is that a tuple $x'$ can be derived by the new Datalog program if and only if $x \in \bP(X)$.
These two rules construct $\bP(X)$ recursively:
The base case \refeqn{revBase} states that the query tuple $\xo \in \bP(X)$.
The recursive case \refeqn{indBase} states that if $x \in \bP(X)$ and a rule $z$ (with some assignment $f$)
was used to produce $x$, then for every source term $s \in z.\bs$ of that rule,
we also have $f(s) \in \bP(X)$.

>>>>

The pruning approach that we present in this paper works on Datalog, a general
language that has been used to express static analyses declaratively \cite{Whaley2007,bravenboer}.  Normally,
these analyses are defined using one monolithic Datalog program which
computes an abstract answer to a query.
For us, it will be convenient to consider the Datalog program, which computes the concrete answer to a query,
as distinct from the abstraction, which is applied on the Datalog program to generate an abstract answer.
This separation will allow us to make theoretical
statements comparing the same Datalog program across different
abstractions.

We first define Datalog (\refsec{datalog}), which describes the
computation of a query (think more in the direction of concrete semantics).
Then, we focus on the abstraction (\refsec{abstractions}), which only interacts
with the Datalog program via the input tuples.
Throughout this section, we will use \reffig{graphExample} as a running example.

@@

A Datalog program takes a set of input tuples and derives new
tuples.  To formalize this computation, we define the notation of a derivation.

A {\em derivation} (of the query $\xo$) with respect to a set of input tuples
$X$ is a sequence $\bx = (x_1, \dots, x_n)$ such that
\begin{enumerate}
\item [(i)] for each $i = 1, \dots, n$, we have $x_i \in X$; or
there exists a set of indices $J$ such that $(J,i)$ satisfies the following conditions:
$j < i$ for each $j \in J$, and
there is a rule $z \in \sZ$ and an assignment $f$
such that $f(z.t) = x_i$ and
$\{ x_j : j \in J \} = \{ f(s) : s \in z.\bs \}$;
\item [(ii)] $x_n = \xo$; and
\item [(iii)] 
for each $j = 1, \dots, n\!-\!1$,
there exists $J$ such that $j \in J$ and an index $i$
such that $(J,i)$ satisfies the conditions in (i).
\end{enumerate}
Define $\bD(X)$ to be the set of all derivations with respect to the input tuples $X$.

% Explain
Condition (i) says that each tuple in a derivation
should either be given as an input tuple ($x_i \in X$)
or be the result of some rule $z \in \sZ$.
Condition (ii) says that $\xo$ was finally produced via some sequence of rule applications.
Condition (iii) says that in the derivation of $x_n$, every tuple is relevant for deriving $\xo$.
%As we will see later, this condition is important because it allow us to
%define the set of irrelevant tuples that can be pruned.

A Datalog program computes $\bD(X)$.
We say that the query $\xo$ is false (proven) if and only if $\bD(X)$ is empty.
Although the query is the ultimate quantity of interest,
the Datalog program can be used to provide more information, which will be useful for pruning.
Specifically, we define $\bP(X)$ to be the subset of the input tuples $X$
which were used in some derivation:
\begin{align}
\label{eqn:Pdef}
\bP(X) \eqdef \{ x \in X : x \in \bx \in \bD(X) \}.
\end{align}
We call $\bP(X)$ the set of {\em relevant input tuples}.  As we will see later,
any tuple not in this set can be safely pruned.
In fact, $\bP(X)$ also tells us whether the query is true or false.
In particular, $\bD(X) = \emptyset$ if and only if $\bP(X) = \emptyset$.
This equivalence suggests that proving and pruning are intimately related;
in some sense, proving the query is just pruning away the query tuple.
In the remainder of the paper, we will make heavy use of $\bP$ as the principal
proving/pruning operation.

@@

We can compute $\bP(X)$ in Datalog by using the Datalog program transformation technique described in 
\cite{liang11minimal}.
This allows us to use off-the-shelf Datalog solvers which have already been optimized
to compute $\bP(X)$.
Specifically, we define a set of new relations $\sR' = \{ r' : r \in \sR \}$.
For a term $t = t.r(t.\ba)$ we let $t' = t.r'(t.\ba)$ be the term that uses the corresponding new relation.
We then add the following new Datalog rules:
\begin{align}
\xo' &\dlogUpdate \xo, \label{eqn:revBase} \\
s'   &\dlogUpdate z.t', z.\bs \quad \text{for each $z \in \sZ$ and $s \in z.\bs$.} \label{eqn:indBase}
\end{align}
The key point is that a tuple $x'$ is derived by the new Datalog program if and only if $x \in \bP(X)$.
These two rules construct $\bP(X)$ recursively:
The base case \refeqn{revBase} states that the query tuple $\xo \in \bP(X)$.
The recursive case \refeqn{indBase} states that if $x \in \bP(X)$ and a rule $z$ (with some assignment $f$)
was used to produce $x$, then for every source term $s \in z.\bs$ of that rule,
we also have $f(s) \in \bP(X)$.
